---
title: "Climate Data Extraction"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/pnwbi/ServerFiles/redcedar")
#knitr::opts_knit$set(root.dir = "/Users/redcedar/ServerFiles/redcedar/") 
```

```{r include=FALSE}
library(tidyverse)
```


## Purpose

The purpose of this markdown document is to detail the methods used to extract climate data for iNat Observations (yes.. windows).

## Data

The data used in the below models are described in the [Data Wrangle folder](./data-wrangle). 

```{r}
data <- read.csv("./data/observations-223649.csv")
```


## Ancillary Climate Data

Climate data then extracted with ClimateNA tool following the below process. Data were downloaded for the iNat GPS locations using the ClimateNA Tool.

ClimateNA version 7.21 - The latest version also supports command line access using system2(exe,"") commands, but you likely need to be running r in windows.

* Climate data extraction process with ClimateNA
  + Convert data into format for climateNA use (see below)
  + In ClimateNA
    + Normal Data
      + Select input file (browse to gps1500 file)
      + Choose 'More Normal Data'
        + Select 'Normal_1991_2020.nrm'
      + Choose 'All variables(265)'
      + Specify output file
    + Decadal Data
      + Select input file (browse to gps1500 file)
      + Choose 'Decadal Data'
        + Select 'Decade_2011_2020.dcd'
      + Choose 'All variables(265)'
      + Specify output file

### Wrangle data into format for climateNA

```{r}
gps <- data[c(1,22,23)] #subset data to only include id and gps coordinates
gps <- rename(gps,lat = latitude) %>% `colnames<-`(c("ID2","lat","long")) %>% mutate(el = ".") #columns were ranamed to match format for ClimateNA tool.
#write.csv(gps,file="./data/gps1500.csv") #named 1500 because data was downloaded after having 1500 observations
```


There were many time series groupings (different intervals where data were averaged) of the downscaled climate variables available. 

* Grouping explored
  + data averaged over 30 year normals (1991-2020)
    + need to check this is actually the time period - I am not sure PRISM had released the latest normals until late 2021.
  + data averaged over last decade (2011-2020)

### Check data

```{r}
normals <- read.csv("./data/gps1500_Normal_1991_2020MSY.csv")
colnames(normals) <- str_c("norm_",colnames(normals)) #change column names - not to be confused with decadal data
normals <- rename(normals, id = norm_ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
str(normals)
```

```{r}
lastdecade <- read.csv("./data/gps1500_Decade_2011_2020MSY.csv")
lastdecade <- rename(lastdecade, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
str(lastdecade)
```