---
title: "iNat Data Analysis"
author: "Joey Hulbert"
date: "4/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/redcedar/ServerFiles/redcedar/") 
```

# Purpose

The purpose of this markdown document is to list the steps completed for analysis of the iNaturalist data. Below we conduct a random forest analysis using only the explanitory variables in the iNaturalist dataset. 

# Data

The data used in the below models are described in the [Data Wrangle folder](./data-wrangle). 

```{r include=FALSE}
library(tidyverse)
library(randomForest)
library(caret)
library(rpart)
library(knitr)
```

```{r}
data <- read.csv("./data/observations-223649.csv")
```

```{r}
data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
data$field.optional...what..other.factors..were.observed. <- as.factor(data$field.optional...what..other.factors..were.observed.)
data$field.tree.canopy.symptoms <- as.factor(data$field.tree.canopy.symptoms)
data$field.optional...slope.position <- as.factor(data$field.optional...slope.position)
data$field.optional...site.type <- as.factor(data$field.optional...site.type)
data$field.optional...site.location.description  <- as.factor(data$field.optional...site.location.description )
data$field.optional...tree.size <-as.factor(data$field.optional...tree.size)
```

```{r}
data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[data$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
```

```{r}
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
data$field.tree.canopy.symptoms[data$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
```

```{r}
data$field.optional...what..other.factors..were.observed.[data$field.optional...what..other.factors..were.observed.=="Fungal Activitiy (mycelial fans, mushrooms at base, or conks on trunk)"] <-"Fungal Activitiy (mycelial fans, bleeding cankers, mushrooms at base, or conks on trunk)"
data$field.optional...what..other.factors..were.observed.[data$field.optional...what..other.factors..were.observed.=="Needle disease (dieback, checking, blight, etc.)"] <- "Needle or leaf disease (dieback, checking, blight, etc.)"
```

```{r}
data$field.optional...slope.position[data$field.optional...slope.position=="Upper 1/3rd of a slope"] <-"Top of slope"
```

```{r}
data$field.optional...site.type[data$field.optional...site.type=="Urban Natural"] <-"Urban"
data$field.optional...site.type[data$field.optional...site.type=="Urban Landscaped"] <-"Urban"
data$field.optional...site.type[data$field.optional...site.type=="Suburban Natural"] <-"Suburban"
data$field.optional...site.type[data$field.optional...site.type=="Suburban Lanscaped"] <-"Suburban"
data$field.optional...site.type[data$field.optional...site.type=="Natural Forest"] <-"Rural"
```

```{r}
data$field.optional...tree.size[data$field.optional...tree.size=="Large"] <- "Large (too big to wrap arms around trunk)"
data$field.optional...tree.size[data$field.optional...tree.size=="Medium"] <- "Medium (can wrap arms around trunk)"
data$field.optional...tree.size[data$field.optional...tree.size=="Small"] <- "Small (can wrap hands around trunk)"
data$field.optional...tree.size[data$field.optional...tree.size=="Very Large"] <- "Very Large (would take many people to wrap arms around trunk)"
```

```{r}
data$field.optional...site.location.description [data$field.optional...site.location.description =="Yard or open park grounds"] <- "Urban yard or open park grounds"
```


```{r}
data <- data %>% droplevels()
```


```{r}
level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Unhealthy", "New Dead Top (red or brown needles still attached)" = "Unhealthy", "Old Dead Top (needles already gone)" = "Unhealthy", "Tree is dead" = "Unhealthy", "Multiple Symptoms (please list in Notes)" = "Unhealthy", "Extra Cone Crop" = "Unhealthy", "Browning Canopy" = "Unhealthy","Branch Dieback or 'Flagging'" = "Unhealthy", "Other (please describe in Notes)" = "Unhealthy", "Yellowing Canopy" = "Unhealthy")

binary <- data
binary$field.tree.canopy.symptoms <- recode_factor(binary$field.tree.canopy.symptoms, !!!level_key)
#levels(binary$field.tree.canopy.symptoms)
binary <- binary %>% filter(field.tree.canopy.symptoms!="Candelabra top or very old spike top (old growth)") %>% droplevels()
binary$field.tree.canopy.symptoms <- as.factor(binary$field.tree.canopy.symptoms)
```


```{r}
binary %>% group_by(field.tree.canopy.symptoms) %>% count()
```

```{r}
names(binary)
binary.factors <- binary %>% select(-c("id"  
,"observed_on_string"                                                                     
,"observed_on"                                                                            
,"time_observed_at"                                                                       
,"time_zone"                                                                              
,"user_id"                                                                                
,"user_login"                                                                             
,"created_at"                                                                             
,"updated_at"
,"quality_grade"                                                                          
,"license"                                                                                
,"url"                                                                                    
,"image_url"                                                                              
,"sound_url"                                                                              
,"tag_list"                                                                               
,"description"                                                                            
,"num_identification_agreements"                                                          
,"num_identification_disagreements"                                                       
,"captive_cultivated"                                                                     
,"oauth_application_id"                                                                   
,"place_guess"                                                                            
,"latitude"                                                                               
,"longitude"                                                                              
,"positional_accuracy"                                                                    
,"private_place_guess"                                                                    
,"private_latitude"                                                                       
,"private_longitude"                                                                      
,"public_positional_accuracy"                                                             
,"geoprivacy"                                                                             
,"taxon_geoprivacy"                                                                       
,"coordinates_obscured"                                                                   
,"positioning_method"                                                                     
,"positioning_device"                                                                     
,"place_town_name"                                                                        
,"place_county_name"                                                                      
,"place_state_name"                                                                       
,"place_country_name"                                                                     
,"place_admin1_name"                                                                      
,"place_admin2_name"                                                                      
,"species_guess"                                                                          
,"scientific_name"
,"common_name"                                                                            
,"iconic_taxon_name"                                                                      
,"taxon_id"
, "field.optional...were.there.any.other.unhealthy.plant.species.on.the.site."             
 ,"field.optional...timing.of.symptoms.estimate"                                           
, "field.optional...estimated.time.spent.to.make.this.observation....of.minutes."          
, "field.optional...can.we.follow.up.with.you."
,"field.percent.canopy.affected...."
,"field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight."
,"field.percent.of.trees..of.same.species..within.sight.that.are.unhealthy"
,"field.notes"
))
```


* Note the following fields were removed because they're not explanitory variables for the below model
  + "field.optional...were.there.any.other.unhealthy.plant.species.on.the.site."             
  + "field.optional...timing.of.symptoms.estimate"                                           
  + "field.optional...estimated.time.spent.to.make.this.observation....of.minutes."          
  + "field.optional...can.we.follow.up.with.you."
  + "field.percent.canopy.affected...."
  + "field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight."
  + "field.percent.of.trees..of.same.species..within.sight.that.are.unhealthy"
  + "field.notes"

#### Full Model

```{r full rf model}
set.seed(71)
rf <- randomForest(field.tree.canopy.symptoms ~ ., data=binary.factors, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9,eval=FALSE}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=binary.factors$field.tree.canopy.symptoms)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_manual(values=c("#7fcdbb","#fe9929")) +
  scale_fill_manual(values=c("#7fcdbb","#fe9929")) +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r}
?varImpPlot
```

```{r fig.height=9,fig.width=15}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```

