---
title: "Redcedar Analyses v1"
author: "Joey Hulbert"
date: "9/19/2021"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r include=FALSE}
#install.packages("tidyverse")
library(tidyverse)
```

## Running questions

* Below are some questions to revisit during further analysis
  + Should we look at data from last decade only rather than last 30 years?
  + How to structure this document
    + climate variable (MAT, MAP)
      + compare across symptoms (filtered?)
      + compare across 'number of unhealthy trees'
      + compare across size class?
      + compare across site type?


## Data Prep

### Import data

Data downloaded manually from iNat. 

```{r}
data <- read.csv("./data/observations-9.19.21.csv")
```

### Prepare dataset for extracting environmental data with climateNA

```{r}
gps <- data[c(1,22,23)]
gps <- rename(gps,lat = latitude) %>% `colnames<-`(c("ID2","lat","long")) %>% mutate(el = ".")
write.csv(gps,file="./data/gps.csv")
```

## 30 Normals 

### Import dataset with 30yr normals data


```{r}
normals <- read.csv("./data/gps_Normal_1991_2020MSY.csv")
normals <- rename(normals, id = ID2)
```


### Merge data

```{r}
combined <- left_join(data,normals,by="id")  %>% filter(MAP >0,na.rm=TRUE)
```

### Clean data 

#### Add consistency to factors
##### Symptoms 
```{r}
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combined <- combined %>% droplevels()
combined$field.tree.canopy.symptoms <- as.factor(combined$field.tree.canopy.symptoms)
levels(combined$field.tree.canopy.symptoms)
```

##### Unhealthy trees
```{r}
combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
combined <- combined %>% droplevels()
combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
levels(combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
```

##### Tree size
```{r}
combined$field.optional...tree.size[combined$field.optional...tree.size == "Large"] <- "Large (too big to wrap arms around trunk)"
combined$field.optional...tree.size[combined$field.optional...tree.size == "Medium"] <- "Medium (can wrap arms around trunk)"
combined$field.optional...tree.size[combined$field.optional...tree.size == "Small"] <- "Small (can wrap hands around trunk)"
combined$field.optional...tree.size[combined$field.optional...tree.size == "Very Large"] <- "Very Large (would take many people to wrap arms around trunk)"
combined <- combined %>% droplevels()
combined$field.optional...tree.size <- as.factor(combined$field.optional...tree.size)
levels(combined$field.optional...tree.size)
```


### Filter data

#### Only include symptoms of interest
```{r}
filtered.symptoms <- combined  %>% filter(field.tree.canopy.symptoms=="Healthy"|field.tree.canopy.symptoms=="Thinning Canopy"|field.tree.canopy.symptoms=="New Dead Top (red or brown needles still attached)"|field.tree.canopy.symptoms=="Old Dead Top (needles already gone)"|field.tree.canopy.symptoms=="Tree is dead") %>% droplevels()
levels(filtered.symptoms$field.tree.canopy.symptoms)
```

### Data Exploration



#### MAT

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```

##### Site Severity

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot()+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="30 Year Normals",x="Symptoms")
```

##### Size Class

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.optional...tree.size,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="30 Year Normals",x="Symptoms")
```


```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+facet_wrap(~field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```


```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +facet_wrap(~field.optional...tree.size) +labs(title="30 Year Normals",x="Symptoms")
```





#### MAP

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```





```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+facet_wrap(~field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +facet_wrap(~field.optional...tree.size) +labs(title="30 Year Normals",x="Symptoms")
```

#### AHM

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,AHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```





```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,AHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+facet_wrap(~field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,AHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +facet_wrap(~field.optional...tree.size) +labs(title="30 Year Normals",x="Symptoms")
```

#### SHM

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```

###### Filtered SHM <200
```{r}
filtered.symptoms.shm <- filtered.symptoms %>% filter(SHM<200) %>% droplevels()
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms", y="Summer Heat-Moisture Index")
```
```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+facet_wrap(~field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +facet_wrap(~field.optional...tree.size) +labs(title="30 Year Normals",x="Symptoms")
```


#### MSP

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms", y="Summer Heat-Moisture Index")
```

#### CMD

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms", y="Summer Heat-Moisture Index")
```


#### Eref

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,Eref,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms")
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,Eref,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="30 Year Normals",x="Symptoms", y="Summer Heat-Moisture Index")
```

## Decade 2011-2020

### Import dataset with Decade 2011-2020


```{r}
lastdecade <- read.csv("./data/gps_Decade_2011_2020MSY.csv")
lastdecade <- rename(lastdecade, id = ID2)
```

### Merge data

```{r}
combined <- left_join(data,lastdecade,by="id")  %>% filter(MAP >0,na.rm=TRUE)
```

### Clean data 

#### Add consistency to factors
##### Symptoms 
```{r}
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combined$field.tree.canopy.symptoms[combined$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combined <- combined %>% droplevels()
combined$field.tree.canopy.symptoms <- as.factor(combined$field.tree.canopy.symptoms)
levels(combined$field.tree.canopy.symptoms)
```

##### Unhealthy trees
```{r}
combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
combined <- combined %>% droplevels()
combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
levels(combined$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
```

##### Tree size
```{r}
combined$field.optional...tree.size[combined$field.optional...tree.size == "Large"] <- "Large (too big to wrap arms around trunk)"
combined$field.optional...tree.size[combined$field.optional...tree.size == "Medium"] <- "Medium (can wrap arms around trunk)"
combined$field.optional...tree.size[combined$field.optional...tree.size == "Small"] <- "Small (can wrap hands around trunk)"
combined$field.optional...tree.size[combined$field.optional...tree.size == "Very Large"] <- "Very Large (would take many people to wrap arms around trunk)"
combined <- combined %>% droplevels()
combined$field.optional...tree.size <- as.factor(combined$field.optional...tree.size)
levels(combined$field.optional...tree.size)
```


### Filter data

#### Only include symptoms of interest
```{r}
filtered.symptoms <- combined  %>% filter(field.tree.canopy.symptoms=="Healthy"|field.tree.canopy.symptoms=="Thinning Canopy"|field.tree.canopy.symptoms=="New Dead Top (red or brown needles still attached)"|field.tree.canopy.symptoms=="Old Dead Top (needles already gone)"|field.tree.canopy.symptoms=="Tree is dead") %>% droplevels()
levels(filtered.symptoms$field.tree.canopy.symptoms)
```

### Data Exploration

#### MAT

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

##### Site Severity

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot()+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="Decade 2011-2020",x="Symptoms")
```

##### Size Class

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.optional...tree.size,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="Decade 2011-2020",x="Symptoms")
```

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+facet_wrap(~field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +facet_wrap(~field.optional...tree.size) +labs(title="Decade 2011-2020",x="Symptoms")
```

#### MAP

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### AHM

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,AHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### SHM

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

###### Filtered SHM <200
```{r}
filtered.symptoms.shm <- filtered.symptoms %>% filter(SHM<200) %>% droplevels()
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms", y="Summer Heat-Moisture Index")
```

#### MSP

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms", y="Summer Heat-Moisture Index")
```

#### CMD

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms", y="Summer Heat-Moisture Index")
```

#### Eref

##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,Eref,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,Eref,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms", y="Summer Heat-Moisture Index")
```

### Large Trees on Severe sites

```{r}
largesevere <- filtered.symptoms %>% filter(field.optional...tree.size=="Large (too big to wrap arms around trunk)" & field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="More than 10" | field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="7-10") %>% droplevels()
```

#### MSP

##### Symptoms

```{r}
ggplot(largesevere,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### CMD

##### Symptoms

```{r}
ggplot(largesevere,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

## TOPO Data


### Import TOPO Data

```{r}
topo <- read.csv("./data/gps plus DEM and TOPO data.csv")
topo <- rename(topo, id = ID2)
```


### Merge data

```{r}
combined2 <- left_join(combined,topo,by="id")
```

### Clean data 

#### Add consistency to factors
##### Symptoms 
```{r}
combined2$field.tree.canopy.symptoms[combined2$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combined2$field.tree.canopy.symptoms[combined2$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combined2$field.tree.canopy.symptoms[combined2$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combined2$field.tree.canopy.symptoms[combined2$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combined2$field.tree.canopy.symptoms[combined2$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combined2 <- combined2 %>% droplevels()
combined2$field.tree.canopy.symptoms <- as.factor(combined2$field.tree.canopy.symptoms)
levels(combined2$field.tree.canopy.symptoms)
```

##### Unhealthy trees
```{r}
combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
combined2 <- combined2 %>% droplevels()
combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
levels(combined2$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
```

##### Tree size
```{r}
combined2$field.optional...tree.size[combined2$field.optional...tree.size == "Large"] <- "Large (too big to wrap arms around trunk)"
combined2$field.optional...tree.size[combined2$field.optional...tree.size == "Medium"] <- "Medium (can wrap arms around trunk)"
combined2$field.optional...tree.size[combined2$field.optional...tree.size == "Small"] <- "Small (can wrap hands around trunk)"
combined2$field.optional...tree.size[combined2$field.optional...tree.size == "Very Large"] <- "Very Large (would take many people to wrap arms around trunk)"
combined2 <- combined2 %>% droplevels()
combined2$field.optional...tree.size <- as.factor(combined2$field.optional...tree.size)
levels(combined2$field.optional...tree.size)
```


### Filter data

#### Only include symptoms of interest
```{r}
filtered.symptoms <- combined2  %>% filter(field.tree.canopy.symptoms=="Healthy"|field.tree.canopy.symptoms=="Thinning Canopy"|field.tree.canopy.symptoms=="New Dead Top (red or brown needles still attached)"|field.tree.canopy.symptoms=="Old Dead Top (needles already gone)"|field.tree.canopy.symptoms=="Tree is dead") %>% droplevels()
levels(filtered.symptoms$field.tree.canopy.symptoms)
```

### Data Exploration

#### Aspect
##### Symptoms

```{r}
names(combined)
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,Aspect,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Aspect (degrees)",x="Symptoms")
```

#### Slope
##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,Slope,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Slope (degrees)",x="Symptoms")
```

#### Elevation
##### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,PNWSRTMDEM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Elevation (meters)",x="Symptoms")
```



### Topo & Climate
#### Slope
##### CMD
```{r}
ggplot(filtered.symptoms,aes(Slope,CMD,color=field.tree.canopy.symptoms))+geom_point()
```

```{r}
ggplot(filtered.symptoms,aes(Slope,CMD,color=field.tree.canopy.symptoms))+geom_smooth()
```

```{r}
ggplot(filtered.symptoms,aes(Slope,CMD,color=field.tree.canopy.symptoms))+geom_smooth()+facet_wrap(~field.tree.canopy.symptoms)
```

#### Elevation
```{r}
names(topo)
ggplot(filtered.symptoms,aes(PNWSRTMDEM,CMD,color=field.tree.canopy.symptoms))+geom_smooth()
```


### Binary Data

```{r}

level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Unhealthy", "New Dead Top (red or brown needles still attached)" = "Unhealthy", "Old Dead Top (needles already gone)" = "Unhealthy", "Tree is dead" = "Unhealthy")

binary <- filtered.symptoms
binary$field.tree.canopy.symptoms <- recode_factor(binary$field.tree.canopy.symptoms, !!!level_key)
levels(binary$field.tree.canopy.symptoms)
```

#### Topo


```{r}
ggplot(binary,aes(PNWSRTMDEM,CMD,color=field.tree.canopy.symptoms))+geom_smooth()
```

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,Slope,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Slope (degrees)",x="Symptoms")
```


#### Climate
### Data Exploration

#### MAT

##### Symptoms

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

##### Site Severity

```{r include=FALSE}
ggplot(binary,aes(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot()+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="Decade 2011-2020",x="Symptoms")
```

##### Size Class

```{r include=FALSE}
ggplot(binary,aes(field.optional...tree.size,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="Decade 2011-2020",x="Symptoms")
```

```{r include=FALSE}
ggplot(binary,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+facet_wrap(~field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

```{r include=FALSE}
ggplot(binary,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +facet_wrap(~field.optional...tree.size) +labs(title="Decade 2011-2020",x="Symptoms")
```

#### MAP

```{r}
names(binary)
```

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,MAP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### AHM

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,AHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### SHM

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

###### Filtered SHM <200
```{r}
binary.shm <- binary %>% filter(SHM<200) %>% droplevels()
```

```{r}
ggplot(binary.shm,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### MSP

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

###### Filtered SHM <200
```{r}
ggplot(binary.shm,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### CMD

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

###### Filtered SHM <200
```{r}
ggplot(binary.shm,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### Eref


```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,Eref,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```
###### Filtered SHM <200

```{r}
ggplot(binary.shm,aes(field.tree.canopy.symptoms,Eref,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

### Large Trees on Severe sites

```{r}
largesevere <- binary %>% filter(field.optional...tree.size=="Large (too big to wrap arms around trunk)" & field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="More than 10" | field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="7-10") %>% droplevels()
```

#### MSP


```{r}
ggplot(largesevere,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### CMD

```{r}
ggplot(largesevere,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```
