---
title: "Climate Data RandomForest Analysis"
author: "Joey Hulbert"
date: "4/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/redcedar/ServerFiles/redcedar/") 
```

```{r include=FALSE}
library(tidyverse)
library(randomForest)
library(caret)
library(rpart)
library(knitr)
```

# Purpose

The purpose of this markdown document is to list the steps we followed for refining the models using climate data only.

# Data

The data used in the below models are described in the [Data Wrangle folder](./data-wrangle). 

## Response variables

**Note the below analysis uses the iNat data with 1510 observations. Amazing!**

* Response variables to consider
  + Number of unhealthy trees
  + Tree canopy symptoms

Below we compare models using canopy symptoms as the response variable

## Explanatory variables

* Explanatory variables included
  + iNat data
    + explanatory variables such as tree size
  + Climate data
    + 30yr normals 1991-2020 (265 variables)
    + last decade  2011-2020 (265 variables)
    
```{r}
data <- read.csv("./data/observations-223649.csv")
```

```{r}
normals <- read.csv("./data/gps1500_Normal_1991_2020MSY.csv")
colnames(normals) <- str_c("norm_",colnames(normals)) #change column names - not to be confused with decadal data
normals <- rename(normals, id = norm_ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
lastdecade <- read.csv("./data/gps1500_Decade_2011_2020MSY.csv")
lastdecade <- rename(lastdecade, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
#combo.normals <- left_join(data,normals,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id")
#combo.decade <- left_join(data,lastdecade,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id") 
combo <- left_join(data,lastdecade,by="id") %>% left_join(.,normals,by="id") 
```

```{r}
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combo <- combo %>% droplevels()
combo$field.tree.canopy.symptoms <- as.factor(combo$field.tree.canopy.symptoms)
#levels(combo$field.tree.canopy.symptoms)
```

```{r}
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
combo <- combo %>% droplevels()
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
#levels(combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
```

```{r}
combo$field.optional...tree.size[combo$field.optional...tree.size == "Large"] <- "Large (too big to wrap arms around trunk)"
combo$field.optional...tree.size[combo$field.optional...tree.size == "Medium"] <- "Medium (can wrap arms around trunk)"
combo$field.optional...tree.size[combo$field.optional...tree.size == "Small"] <- "Small (can wrap hands around trunk)"
combo$field.optional...tree.size[combo$field.optional...tree.size == "Very Large"] <- "Very Large (would take many people to wrap arms around trunk)"
combo <- combo %>% droplevels()
combo$field.optional...tree.size <- as.factor(combo$field.optional...tree.size)
#levels(combo$field.optional...tree.size)
```

# Data wrangling

There are multiple methods to group the response variables deepening on desired resolution or fineness of the model. 

* Response grouping options
  + Original (no change)
  + Filtered Symptoms
    + We could drop trees outside of main symptom classes
       + Note loss of data outside of categories
  + Reclassified symptoms
    + Here we can created an 'other' variable to represent trees with the below symptoms
        + extra cone crop
        + yellowing canopy
    + We can also group trees with 'old dead tops' and 'new dead tops' into a 'dead top' category
  + Binary (healthy/unhealthy)

For now, we can move forward with the binary response grouping because it is the broadest and presumably the easiest for the model to classify with. 


All tree health categories

```{r}
combo %>% group_by(field.tree.canopy.symptoms) %>% count()
```

We also need to filter the data to only include response and explanatory variables we're interested in. For example, whether a sound clip was included in the iNat data is not important. 

We also need to remove other response variables like "field.percent.canopy.affected...." so it is not used as a predictor for tree health. 

> Note it might be interesting to know if the user was an important factor in predicting if the tree is healthy/unhealthy. 

There are also a number of factors that should probably be removed because they may be biasing the data. For example, only trees with the 'other factor' question may only be answered for unhealthy trees. We need to think about this a bit more. 

```{r}
combo.filtered <- combo %>% select(-c("field.optional...site.type",
                                      "field.optional...site.hydrology",
                                      "field.optional...site.location.description",
                                      "field.optional...site.area.disturbance.level",
                                      "field.optional...tree.size",
  "field.optional...slope.position",
  "field.optional...did.the.tree.have.heat.damage",
,"field.percent.canopy.affected...."
,"field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight."
,"field.percent.of.trees..of.same.species..within.sight.that.are.unhealthy"
,"field.optional...access.to.water"
,"id"  
,"observed_on_string"                                                                     
,"observed_on"                                                                            
,"time_observed_at"                                                                       
,"time_zone"                                                                              
,"user_id"                                                                                
,"user_login"                                                                             
,"created_at"                                                                             
,"updated_at"
,"quality_grade"                                                                          
,"license"                                                                                
,"url"                                                                                    
,"image_url"                                                                              
,"sound_url"                                                                              
,"tag_list"                                                                               
,"description"                                                                            
,"num_identification_agreements"                                                          
,"num_identification_disagreements"                                                       
,"captive_cultivated"                                                                     
,"oauth_application_id"                                                                   
,"place_guess"                                                                            
,"latitude"                                                                               
,"longitude"                                                                              
,"positional_accuracy"                                                                    
,"private_place_guess"                                                                    
,"private_latitude"                                                                       
,"private_longitude"                                                                      
,"public_positional_accuracy"                                                             
,"geoprivacy"                                                                             
,"taxon_geoprivacy"                                                                       
,"coordinates_obscured"                                                                   
,"positioning_method"                                                                     
,"positioning_device"                                                                     
,"place_town_name"                                                                        
,"place_county_name"                                                                      
,"place_state_name"                                                                       
,"place_country_name"                                                                     
,"place_admin1_name"                                                                      
,"place_admin2_name"                                                                      
,"species_guess"                                                                          
,"scientific_name"
,"common_name"                                                                            
,"iconic_taxon_name"                                                                      
,"taxon_id"          
,"field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage."
,"field.optional...what..other.factors..were.observed." 
, "field.optional...were.there.any.other.unhealthy.plant.species.on.the.site."             
 ,"field.optional...timing.of.symptoms.estimate"                                           
, "field.optional...estimated.time.spent.to.make.this.observation....of.minutes."          
, "field.optional...can.we.follow.up.with.you."
,"X" 
,"Latitude"                                    
,"Longitude"
,"field.notes"
,"norm_X"                                                                                   
,"norm_Latitude"                                                                            
,"norm_Longitude"                                                                           
,"norm_Elevation"
))
# combo.filtered <- combo.filtered[c(45,50:59,68:332,337:341,355:394,396:503,510:774)] 
```

Remove variables with variables that have near zero standard deviations (entire column is same value)

```{r include=FALSE}
#nearZeroVar(combo.filtered)
combo.filtered.nearzerovar <- combo.filtered[,-nearZeroVar(combo.filtered)]
#write.csv(combo.decade.filtered,file="./data/troubleshoot.csv")
```

```{r}
combo.filtered.nearzerovar <-as.data.frame(unclass(combo.filtered.nearzerovar),stringsAsFactors=TRUE) #change all chars to factors
```


## Group response variables to binary values

```{r}
level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Unhealthy", "New Dead Top (red or brown needles still attached)" = "Unhealthy", "Old Dead Top (needles already gone)" = "Unhealthy", "Tree is dead" = "Unhealthy", "Multiple Symptoms (please list in Notes)" = "Unhealthy", "Extra Cone Crop" = "Unhealthy", "Browning Canopy" = "Unhealthy","Branch Dieback or 'Flagging'" = "Unhealthy", "Other (please describe in Notes)" = "Unhealthy", "Yellowing Canopy" = "Unhealthy")

binary <- combo.filtered.nearzerovar
binary$field.tree.canopy.symptoms <- recode_factor(binary$field.tree.canopy.symptoms, !!!level_key)
#levels(binary$field.tree.canopy.symptoms)
binary <- binary %>% filter(field.tree.canopy.symptoms!="Candelabra top or very old spike top (old growth)") %>% droplevels()
binary$field.tree.canopy.symptoms <- as.factor(binary$field.tree.canopy.symptoms)
```

Binary tree health categories

```{r}
binary %>% group_by(field.tree.canopy.symptoms) %>% count()
```

# Approach

* General approach
  + build full model using binary response grouping variable
    + binary response 
  + refine model to remove redundant explanatory variables
    
* Approach notes
  + We need to train/test models to calculate correctly predicted presences (inverse of commission errors) and correctly predicted absences (inverse of omission errors)
  + we may want to evaluate the climate parameters seperately
    + we lose >300 observations when we filter out data to only include those with soil
    + what happens if we build a model without soils data (without filtering to WA and OR)

# Models

* Random forest model criteria
  + nearzero variation columns were removed
  + Data were imputed prior to model building. 
  + number of splits was not defined to allow automatic selection,
  + ntree = 2001, importance = TRUE, proximity = TRUE, na.action = na.omit
  + seed was set for reproducability with: set.seed(71)

## Comparing full and reduced datasets of explanatory variables  

### Symtpom response variable

 * Datasets compared below
    + Full model = all climate variables retained. 
    + Monthless model = filtered out the climate parameters for individual months
    + Normal only, monthless model = removed decadal data

```{r}
monthless.binary <- binary[-c(9:155,232:378)]
```

```{r}
normal.monthless.binary <- binary[-c(9:378)]
```


#### Full Model

```{r full rf model}
set.seed(71)
rf <- randomForest(field.tree.canopy.symptoms ~ ., data=binary, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=binary$field.tree.canopy.symptoms)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_manual(values=c("#7fcdbb","#fe9929")) +
  scale_fill_manual(values=c("#7fcdbb","#fe9929")) +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```

#### Monthless Model

```{r monthless rf model}
set.seed(71)
rf <- randomForest(field.tree.canopy.symptoms ~ ., data=monthless.binary, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
?randomForest
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=monthless.binary$field.tree.canopy.symptoms)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_manual(values=c("#7fcdbb","#fe9929")) +
  scale_fill_manual(values=c("#7fcdbb","#fe9929")) +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```

#### Normal, Monthless Model


```{r normal monthless model}
set.seed(71)
rf <- randomForest(field.tree.canopy.symptoms ~ ., data=normal.monthless.binary, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
?randomForest
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=normal.monthless.binary$field.tree.canopy.symptoms)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_manual(values=c("#7fcdbb","#fe9929")) +
  scale_fill_manual(values=c("#7fcdbb","#fe9929")) +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```


### Number of unhealthy trees response variable

```{r}
combo.filtered <- combo %>% select(-c("field.optional...site.type",
                                      "field.optional...site.hydrology",
                                      "field.optional...site.location.description",
                                      "field.optional...site.area.disturbance.level",
                                      "field.optional...tree.size",
  "field.optional...slope.position",
  "field.optional...did.the.tree.have.heat.damage",
  "field.tree.canopy.symptoms",
,"field.percent.canopy.affected...."
,"field.percent.of.trees..of.same.species..within.sight.that.are.unhealthy"
,"field.optional...access.to.water"
,"id"  
,"observed_on_string"                                                                     
,"observed_on"                                                                            
,"time_observed_at"                                                                       
,"time_zone"                                                                              
,"user_id"                                                                                
,"user_login"                                                                             
,"created_at"                                                                             
,"updated_at"
,"quality_grade"                                                                          
,"license"                                                                                
,"url"                                                                                    
,"image_url"                                                                              
,"sound_url"                                                                              
,"tag_list"                                                                               
,"description"                                                                            
,"num_identification_agreements"                                                          
,"num_identification_disagreements"                                                       
,"captive_cultivated"                                                                     
,"oauth_application_id"                                                                   
,"place_guess"                                                                            
,"latitude"                                                                               
,"longitude"                                                                              
,"positional_accuracy"                                                                    
,"private_place_guess"                                                                    
,"private_latitude"                                                                       
,"private_longitude"                                                                      
,"public_positional_accuracy"                                                             
,"geoprivacy"                                                                             
,"taxon_geoprivacy"                                                                       
,"coordinates_obscured"                                                                   
,"positioning_method"                                                                     
,"positioning_device"                                                                     
,"place_town_name"                                                                        
,"place_county_name"                                                                      
,"place_state_name"                                                                       
,"place_country_name"                                                                     
,"place_admin1_name"                                                                      
,"place_admin2_name"                                                                      
,"species_guess"                                                                          
,"scientific_name"
,"common_name"                                                                            
,"iconic_taxon_name"                                                                      
,"taxon_id"          
,"field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage."
,"field.optional...what..other.factors..were.observed." 
, "field.optional...were.there.any.other.unhealthy.plant.species.on.the.site."             
 ,"field.optional...timing.of.symptoms.estimate"                                           
, "field.optional...estimated.time.spent.to.make.this.observation....of.minutes."          
, "field.optional...can.we.follow.up.with.you."
,"X" 
,"Latitude"                                    
,"Longitude"
,"field.notes"
,"norm_X"                                                                                   
,"norm_Latitude"                                                                            
,"norm_Longitude"                                                                           
,"norm_Elevation"
))
# combo.filtered <- combo.filtered[c(45,50:59,68:332,337:341,355:394,396:503,510:774)] 
```

Remove variables with variables that have near zero standard deviations (entire column is same value)

```{r include=FALSE}
#nearZeroVar(combo.filtered)
combo.filtered.nearzerovar <- combo.filtered[,-nearZeroVar(combo.filtered)]
#write.csv(combo.decade.filtered,file="./data/troubleshoot.csv")
```

```{r}
combo.filtered.nearzerovar <-as.data.frame(unclass(combo.filtered.nearzerovar),stringsAsFactors=TRUE) #change all chars to factors
```

```{r}
monthless.combo.filtered.nearzerovar <- combo.filtered.nearzerovar[-c(9:155,232:378)]
```

```{r}
normal.monthless.combo.filtered.nearzerovar <- combo.filtered.nearzerovar[-c(9:378)]
```

#### Full Model

```{r number trees full rf model}
set.seed(71)
rf <- randomForest(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. ~ ., data=combo.filtered.nearzerovar, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=combo.filtered.nearzerovar$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_discrete() +
  scale_fill_discrete() +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```

#### Monthless Model

```{r number trees monthless rf model}
set.seed(71)
rf <- randomForest(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. ~ ., data=monthless.combo.filtered.nearzerovar , ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
?randomForest
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=monthless.combo.filtered.nearzerovar$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_discrete() +
  scale_fill_discrete() +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```

#### Normal, Monthless Model


```{r number trees normal monthless model}
set.seed(71)
rf <- randomForest(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. ~ ., data=normal.monthless.combo.filtered.nearzerovar, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
?randomForest
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=normal.monthless.combo.filtered.nearzerovar$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_discrete() +
  scale_fill_discrete() +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```