---
title: "Climate Data RandomForest Analysis - unhealthy tree number response"
author: "Joey Hulbert"
date: "4/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/redcedar/ServerFiles/redcedar/") 
```

```{r include=FALSE}
library(tidyverse)
library(randomForest)
library(caret)
library(rpart)
library(knitr)
```

# Purpose

The purpose of this markdown document is to list the steps we followed for refining the models using climate data only.

# Data

The data used in the below models are described in the [Data Wrangle folder](./data-wrangle). 

## Response variables

**Note the below analysis uses the iNat data with 1510 observations. Amazing!**

* Response variables to consider
  + Number of unhealthy trees

Below we compare models using canopy symptoms as the response variable

## Explanatory variables

* Explanatory variables included
  + Climate data
    + 30yr normals 1991-2020 (265 variables)
    
```{r}
data <- read.csv("./data/observations-223649.csv")
#data.filtered <- data %>% select(c("id","field.tree.canopy.symptoms")) - remove all data except response variable of interest
```

```{r}
normals <- read.csv("./data/gps1500_Normal_1991_2020MSY.csv",na.string ="-9999")
colnames(normals) <- str_c("norm_",colnames(normals)) #change column names - not to be confused with decadal data
normals <- rename(normals, id = norm_ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
lastdecade <- read.csv("./data/gps1500_Decade_2011_2020MSY.csv",na.string ="-9999")
lastdecade <- rename(lastdecade, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

> Note that we included na.string="-9999" when reading data. 

```{r}
#combo.normals <- left_join(data,normals,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id")
#combo.decade <- left_join(data,lastdecade,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id") 
combo <- left_join(data,lastdecade,by="id") %>% left_join(.,normals,by="id") 
```

```{r}
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combo <- combo %>% droplevels()
combo$field.tree.canopy.symptoms <- as.factor(combo$field.tree.canopy.symptoms)
#levels(combo$field.tree.canopy.symptoms)
```

```{r}
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
combo <- combo %>% droplevels()
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
#levels(combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
```

# Data wrangling

There are multiple methods to group the response variables deepening on desired resolution or fineness of the model. 

We also need to filter the data to only include response and explanatory variables we're interested in. For example, whether a sound clip was included in the iNat data is not important. 
We also need to remove other response variables like "field.percent.canopy.affected...." so it is not used as a predictor for tree health. 

> Note it might be interesting to know if the user was an important factor in predicting if the tree is healthy/unhealthy. 

There are also a number of factors that should probably be removed because they may be biasing the data. For example, only trees with the 'other factor' question may only be answered for unhealthy trees. We need to think about this a bit more. 

First remove iNat factors not useful as explanatory variables (e.g. observed_on)

```{r}
combo.filtered <- combo %>% select(-c("field.optional...site.type",
                                      "field.optional...site.hydrology",
                                      "field.optional...site.location.description",
                                      "field.optional...site.area.disturbance.level",
                                      "field.optional...tree.size",
  "field.optional...slope.position",
  "field.optional...did.the.tree.have.heat.damage"
,"field.percent.canopy.affected...."
,"field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight."
,"field.percent.of.trees..of.same.species..within.sight.that.are.unhealthy"
,"field.optional...access.to.water"
,"id"  
,"observed_on_string"                                                                     
,"observed_on"                                                                            
,"time_observed_at"                                                                       
,"time_zone"                                                                              
,"user_id"                                                                                
,"user_login"                                                                             
,"created_at"                                                                             
,"updated_at"
,"quality_grade"                                                                          
,"license"                                                                                
,"url"                                                                                    
,"image_url"                                                                              
,"sound_url"                                                                              
,"tag_list"                                                                               
,"description"                                                                            
,"num_identification_agreements"                                                          
,"num_identification_disagreements"                                                       
,"captive_cultivated"                                                                     
,"oauth_application_id"                                                                   
,"place_guess"                                                                            
,"latitude"                                                                               
,"longitude"                                                                              
,"positional_accuracy"                                                                    
,"private_place_guess"                                                                    
,"private_latitude"                                                                       
,"private_longitude"                                                                      
,"public_positional_accuracy"                                                             
,"geoprivacy"                                                                             
,"taxon_geoprivacy"                                                                       
,"coordinates_obscured"                                                                   
,"positioning_method"                                                                     
,"positioning_device"                                                                     
,"place_town_name"                                                                        
,"place_county_name"                                                                      
,"place_state_name"                                                                       
,"place_country_name"                                                                     
,"place_admin1_name"                                                                      
,"place_admin2_name"                                                                      
,"species_guess"                                                                          
,"scientific_name"
,"common_name"                                                                            
,"iconic_taxon_name"                                                                      
,"taxon_id"          
,"field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage."
,"field.optional...what..other.factors..were.observed." 
, "field.optional...were.there.any.other.unhealthy.plant.species.on.the.site."             
 ,"field.optional...timing.of.symptoms.estimate"                                           
, "field.optional...estimated.time.spent.to.make.this.observation....of.minutes."          
, "field.optional...can.we.follow.up.with.you."
,"field.notes"
))
# combo.filtered <- combo.filtered[c(45,50:59,68:332,337:341,355:394,396:503,510:774)] 
```

Then, remove specific climate variables not useful as explanatory variables (e.g. norm_Latitutde)
```{r}
combo.filtered <- combo.filtered %>% select(-c(
"X" 
,"Latitude"                                    
,"Longitude"
,"norm_X"                                                                                   
,"norm_Latitude"                                                                            
,"norm_Longitude"                                                                           
,"norm_Elevation"))
```


Remove variables with variables that have near zero standard deviations (entire column is same value)

```{r include=FALSE}
#nearZeroVar(combo.filtered)
combo.filtered.nearzerovar <- combo.filtered[,-nearZeroVar(combo.filtered)]
#write.csv(combo.decade.filtered,file="./data/troubleshoot.csv")
```

```{r}
combo.filtered.nearzerovar <-as.data.frame(unclass(combo.filtered.nearzerovar),stringsAsFactors=TRUE) #change all chars to factors
```

# Approach

* General approach
  + build full model using binary response grouping variable
    + binary response 
  + refine model to remove redundant explanatory variables
  

# Models

* Random forest model criteria
  + nearzero variation columns were removed
  + Data were imputed prior to model building. 
  + number of splits was not defined to allow automatic selection,
  + ntree = 2001, importance = TRUE, proximity = TRUE, na.action = na.omit
  + seed was set for reproducability with: set.seed(71)

## Comparing full and reduced datasets of explanatory variables  

### Number of unhealthy trees response variable

```{r}
combo.filtered <- combo %>% select(-c("field.optional...site.type",
                                      "field.optional...site.hydrology",
                                      "field.optional...site.location.description",
                                      "field.optional...site.area.disturbance.level",
                                      "field.optional...tree.size",
  "field.optional...slope.position",
  "field.optional...did.the.tree.have.heat.damage",
  "field.tree.canopy.symptoms",
"field.percent.canopy.affected...."
,"field.percent.of.trees..of.same.species..within.sight.that.are.unhealthy"
,"field.optional...access.to.water"
,"id"  
,"observed_on_string"                                                                     
,"observed_on"                                                                            
,"time_observed_at"                                                                       
,"time_zone"                                                                              
,"user_id"                                                                                
,"user_login"                                                                             
,"created_at"                                                                             
,"updated_at"
,"quality_grade"                                                                          
,"license"                                                                                
,"url"                                                                                    
,"image_url"                                                                              
,"sound_url"                                                                              
,"tag_list"                                                                               
,"description"                                                                            
,"num_identification_agreements"                                                          
,"num_identification_disagreements"                                                       
,"captive_cultivated"                                                                     
,"oauth_application_id"                                                                   
,"place_guess"                                                                            
,"latitude"                                                                               
,"longitude"                                                                              
,"positional_accuracy"                                                                    
,"private_place_guess"                                                                    
,"private_latitude"                                                                       
,"private_longitude"                                                                      
,"public_positional_accuracy"                                                             
,"geoprivacy"                                                                             
,"taxon_geoprivacy"                                                                       
,"coordinates_obscured"                                                                   
,"positioning_method"                                                                     
,"positioning_device"                                                                     
,"place_town_name"                                                                        
,"place_county_name"                                                                      
,"place_state_name"                                                                       
,"place_country_name"                                                                     
,"place_admin1_name"                                                                      
,"place_admin2_name"                                                                      
,"species_guess"                                                                          
,"scientific_name"
,"common_name"                                                                            
,"iconic_taxon_name"                                                                      
,"taxon_id"          
,"field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage."
,"field.optional...what..other.factors..were.observed." 
, "field.optional...were.there.any.other.unhealthy.plant.species.on.the.site."             
 ,"field.optional...timing.of.symptoms.estimate"                                           
, "field.optional...estimated.time.spent.to.make.this.observation....of.minutes."          
, "field.optional...can.we.follow.up.with.you."
,"X" 
,"Latitude"                                    
,"Longitude"
,"field.notes"
,"norm_X"                                                                                   
,"norm_Latitude"                                                                            
,"norm_Longitude"                                                                           
,"norm_Elevation"
))
# combo.filtered <- combo.filtered[c(45,50:59,68:332,337:341,355:394,396:503,510:774)] 
```

Remove variables with variables that have near zero standard deviations (entire column is same value)

```{r include=FALSE}
#nearZeroVar(combo.filtered)
combo.filtered.nearzerovar <- combo.filtered[,-nearZeroVar(combo.filtered)]
#write.csv(combo.decade.filtered,file="./data/troubleshoot.csv")
```

```{r}
combo.filtered.nearzerovar <- na.omit(combo.filtered.nearzerovar)
```

```{r}
combo.filtered.nearzerovar <-as.data.frame(unclass(combo.filtered.nearzerovar),stringsAsFactors=TRUE) #change all chars to factors
```

```{r}
monthless.combo.filtered.nearzerovar <- combo.filtered.nearzerovar[-c(9:155,232:378)]
```

```{r}
normal.monthless.combo.filtered.nearzerovar <- combo.filtered.nearzerovar[-c(9:378)]
```



#### Normal, Monthless Model


```{r number trees normal monthless model}
set.seed(71)
rf <- randomForest(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. ~ ., data=normal.monthless.combo.filtered.nearzerovar, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=normal.monthless.combo.filtered.nearzerovar$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_discrete() +
  scale_fill_discrete() +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```


