---
title: "Topo Data Extraction - ArcGIS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this Markdown Document is to record the methods for extracting topo data for the 1500 redcedar GPS points using ArcGIS.

**Note the below data can only be extracted for trees within the US. :( **

## ArcGIS Toolbox

Topo data were extracted for each point using the Geomorphometry and Gradient Metrics Toolbox available [here](https://www.arcgis.com/home/item.html?id=63ffcecf3b2a45bf99a84cdaedefaccf). 

The toolbox was developed by Jeffrey Evans. More information is available at: 
https://evansmurphy.wixsite.com/evansspatial/arcgis-gradient-metrics-toolbox

Evans JS, Oakleaf J, Cushman SA (2014) An ArcGIS Toolbox for Surface Gradient and Geomorphometric Modeling, version 2.0-0. URL: https://github.com/jeffreyevans/GradientMetrics Accessed: `r format(Sys.time(), '%d %B, %Y')`.

A manuscript about the package is available [here](https://www.sciencedirect.com/science/article/pii/S0098300415300613). 

## Explanitory variables

This package calculates neat data for DEM rasters that can be useful as explanitory variables for the response of tree health. For example, this dataset includes calculations for heat load index and transformations for aspect. 

* Variables
  + CTI - similar to TWI
  + Slope Position
  + Heat Load Index - slope, aspect and latitude
  + 2nd derivitive slope - concave vs convexity? 
  + Transformed Aspect 
  

> Note you will need to extract data to points for each of the calculated rasters above. We will also want some data directly extracted from DEMs


## Data

The data used in the below models are described in the [Data Wrangle folder](./data-wrangle). 

### Download DEMs


#### Earth Explorer Method (errors)

https://earthexplorer.usgs.gov/ - Note these rasters may be too large or have too poor of resolution for use with the geomorphometric toolbox. 

We kept getting errors while trying to create HLI rasters, so we decided to try a DEMs with higher resolution and smaller areas. 

* Earth Explorer (https://earthexplorer.usgs.gov/)
  + Search Criteria
    + Circle, selected seattle as center point
    + 600 km radius
  + Data sets
    + Digital Elevation
      + GTOP030
  + Results
    + Downloaded two DEMS
      + Entity ID: GT30W140N40
      + Entity ID: GT30W140N90
      
      
#### National Map Method (recommended)

* National Map (https://apps.nationalmap.gov/downloader/#/)
  + Define Area of Interest
    + Selectable Polygon
      + US State or Territory
        + Washington (repeat below with Oregon)
  + Search Products
    + Elevational Products (3DEP)
      + 1/3 arc-second DEM
        + Current
        + 1x1 degree
        + GeoTiff
  + Add all relevant products to cart (repeat with Oregon)
  + Download Data
    + follow instructions for batch download using uGet
    + Note each tile is about 500MBs

DEMs downloaded for oregon and Washington (see r markdone script for list of urls to tifs)

```{r include=FALSE, eval=FALSE}
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w116/USGS_13_n45w116_20220309.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w117/USGS_13_n45w117_20220309.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w116/USGS_13_n46w116_20220309.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w117/USGS_13_n46w117_20220309.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w119/USGS_13_n46w119_20220309.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w120/USGS_13_n46w120_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w121/USGS_13_n46w121_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w122/USGS_13_n46w122_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w123/USGS_13_n46w123_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w118/USGS_13_n47w118_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w119/USGS_13_n47w119_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w120/USGS_13_n47w120_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w121/USGS_13_n47w121_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w122/USGS_13_n47w122_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w123/USGS_13_n47w123_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w124/USGS_13_n47w124_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n48w118/USGS_13_n48w118_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n48w119/USGS_13_n48w119_20211004.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n48w122/USGS_13_n48w122_20210615.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n48w123/USGS_13_n48w123_20210615.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n49w118/USGS_13_n49w118_20211129.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w125/USGS_13_n47w125_20200213.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n48w124/USGS_13_n48w124_20200109.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n48w125/USGS_13_n48w125_20200109.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n49w123/USGS_13_n49w123_20200107.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n49w124/USGS_13_n49w124_20200109.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n49w125/USGS_13_n49w125_20200110.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w116/USGS_13_n44w116_20190614.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w117/USGS_13_n44w117_20190614.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w118/USGS_13_n44w118_20170417.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w119/USGS_13_n44w119_20191127.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w120/USGS_13_n44w120_20191127.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w121/USGS_13_n44w121_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w122/USGS_13_n44w122_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w123/USGS_13_n44w123_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w124/USGS_13_n44w124_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n44w125/USGS_13_n44w125_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w118/USGS_13_n45w118_20181128.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w119/USGS_13_n45w119_20180501.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w120/USGS_13_n45w120_20180501.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w121/USGS_13_n45w121_20180501.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w122/USGS_13_n45w122_20140718.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w123/USGS_13_n45w123_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w124/USGS_13_n45w124_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n45w125/USGS_13_n45w125_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w118/USGS_13_n46w118_20190515.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w124/USGS_13_n46w124_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n46w125/USGS_13_n46w125_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w116/USGS_13_n47w116_20190918.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n47w117/USGS_13_n47w117_20190918.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w125/USGS_13_n43w125_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w122/USGS_13_n43w122_20191125.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w121/USGS_13_n43w121_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w120/USGS_13_n43w120_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w119/USGS_13_n43w119_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w118/USGS_13_n43w118_20170417.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w117/USGS_13_n43w117_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n42w121/USGS_13_n42w121_20190313.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n42w125/USGS_13_n42w125_20200122.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w124/USGS_13_n43w124_20210623.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n43w123/USGS_13_n43w123_20210623.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n42w124/USGS_13_n42w124_20210623.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n42w123/USGS_13_n42w123_20210623.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n42w122/USGS_13_n42w122_20210624.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n48w120/USGS_13_n48w120_20161028.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n48w121/USGS_13_n48w121_20130911.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n49w119/USGS_13_n49w119_20161028.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n49w120/USGS_13_n49w120_20161028.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n49w121/USGS_13_n49w121_20180202.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n49w122/USGS_13_n49w122_20180202.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n50w122/USGS_13_n50w122_20180202.tif
https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/13/TIFF/historical/n50w123/USGS_13_n50w123_20180202.tif
```

## Methods

* Process
  + Add iNat observations
    + Map tab -> Add Data -> Browse to below folder
    + C:\Users\pnwbi\ServerFiles\redcedar\data
          + Add CSV with iNat GPS - Add GPS points wrangled from data downloaded from iNaturalist project
          + Note ArcGIS Pro was doing a weird thing where after opening gps csv, it wouldn't recognize lat/long columns the next time I tried to add it as a layer.
            + Copy original GPS file as a new file specifically for use in ARCGIS
          + If you add csv as table with 'add data' button in Map Tab of ArcGIS Pro, you can then right click the layer and select 'display XY data'
            + It will add a layer with your points
  + Add DEM Tiles
    + Add tiles (.tif) downloaded from National Map
      + use 'Add Data' button and browse to DEM .tifs
      + note this took a while to add the data when you add them all at once. 
  + Calculate HLI (Heat Load Index) for each tile
    + Geoprocessing -> Heat Load Index (search for tool or find in catalog/toolboxes/geomporph..)
      + Select DEM (see list below)  - one at a time.
      + Add 'HLI_' to front of output_raster name. No need to add file type designation, just leave blank (e.g. HLI_USGS_13_n48w124_20200109)
  + Repeat for the following measurements (Descriptions Below)
    + CTI
    + TRASP
    + SDS
    + IMI
    + SP - Slope Position
      + SP was calculated using default settings (Rectangle, Width=3, Height=3, Units type = Cell)
      + Some slope position rasters were funky looking without variation in tile.
    + LF - Landform
  + Merge Rasters (search for tool - Mosaic to New Raster) - followed instructions at [How To: Merge multiple raster datasets into a new raster dataset in ArcGIS Pro](https://support.esri.com/en/technical-article/000022861) and [How To: Merge multiple raster datasets into a new raster dataset in ArcMap](https://support.esri.com/en/Technical-Article/000015258).
    + click the down arrow for 'merge many'
    + select input rasters you want to merge
    + Select output location (same geodatabase)
    + specify name
    + specify pixel type - use same pixel type as input rasters (32 for CTI rasters)
      + right click on input rasters in console, select properties, click raster info, look for pixel type and number of bands
      + Note we chose 32 pixels 'unspecified'
    + specify number of bands (1 for CTI rasters)

> Note, I am not sure if it is faster to merge all rasters, then try to extract from rasters for all points. Or whether it is faster to subset the points into each tile, then extract form each raster and tile. 
  
> Note I had to add () after the 'print' command in scripts for CTI and IMI (docs/arc/toolbox/geomor/...)

    
  * DEM tiles used, overlaying points (tiles for some ID, MT, and CA points not calculated at this time)
    + USGS_13_n49w125_20200110
    + USGS_13_n49w124_20200109
    + USGS_13_n49w123_20200107
    + USGS_13_n49w122_20180202
    + USGS_13_n49w121_20180202
    + USGS_13_n49w119_20161028
    + USGS_13_n49w118_20211129
    + USGS_13_n48w125_20200109
    + USGS_13_n48w124_20200109
    + USGS_13_n48w123_20210615
    + USGS_13_n48w122_20210615
    + USGS_13_n48w121_20130911
    + USGS_13_n48w118_20211129
    + USGS_13_n47w124_20211129
    + USGS_13_n47w123_20211129
    + USGS_13_n47w122_20211129
    + USGS_13_n47w118_20211129
    + USGS_13_n47w117_20190918
    + USGS_13_n47w116_20190918
    + USGS_13_n46w125_20130911
    + USGS_13_n46w124_20130911
    + USGS_13_n46w123_20211129
    + USGS_13_n46w122_20211129
    + USGS_13_n45w124_20130911
    + USGS_13_n45w123_20130911
    + USGS_13_n45w122_20140718
    + USGS_13_n44w125_20130911
    + USGS_13_n44w124_20130911
    + USGS_13_n44w123_20130911
    
  


* Variables included are described by -[Evans et al. 2014](https://github.com/jeffreyevans/GradientMetrics) as follows.
    + Temperature & Moisture
      + HLI - Heat Load Index - "A southwest facing slope should have warmer temperatures than a southeast facing slope, even though the amount of solar radiation they recieve is equivalent. The McCune and Keon (2002) method accounts for this by "folding" the aspect so that the highest values are southwest and the lowest values are northeast. Additionally, this method account for steepness of slope, which is not addressed in most other aspect rescaling equations." 
        + McCune and Keon (2002). McCune, Bruce and Dylan Keon, 2002. Equations for potential annual direct incident radiation and heat load index. Journal of Vegetation Science. 13:603-606.
      + CTI - Compound Topographic Index - "steady state wetness index. The CTI is a function of both the slope and the upstream contributing area per unit width orthigonal to the flow direction" -[Evans et al. 2014](https://github.com/jeffreyevans/GradientMetrics)
      + TRASP - Slope/Aspect Transformation - Topographic radiation aspect index (Robert & Cooper 1989) - "Circular aspect is transformed to assign a value of zero to land oriented in a north-northeast direction, (typically the coolest and wettest orientation), and a value of one on the hnotter, dryer south-southwesterly slopes. The result is a continuous variable between 0-1).
        + Roberts. D. W., and Cooper, S. V., 1989. Concepts and techniques of vegetation mapping. In Land Classifications Based on Vegetation: Applications for Resource Management. USDA Forest Service GTR INT-257, Ogden, UT, pp 90-96
      + SDS - 2nd Derivative Slope - " Calculates 2nd derivative of slope" 
        + the rate of the rate of change in slope? concave or convex? 
      + IMI - Integrated Moisture Index - "Calculates an integrated moisture index (Iverson et al. 1997)" 
        + [Iverson, Louis R., et al. "A GIS-derived integrated moisture index to predict forest composition and productivity of Ohio forests (USA)." Landscape Ecology 12.5 (1997): 331-348.](https://www.srs.fs.usda.gov/pubs/11583)    
    + Surface Texture
      + Slope Position - "calculates scalable slope position by subtracting a focalmean raster from the original elevation raster. Surface/Area Ratio - (float) The Berry (2002) method for surface/area ratio." 
        + [Berry, J. K. 2002. Use surface area for realistic calculations. Geoworld 15(9): 20–1.]()
      + Landform- "Concavity/convexity landform index (Bolstad’s variant)." 
  

  
  + Use DEMS to generate HLIs for each tile
  + Merge HLI Rasters
  + Extract HLI data for points in rasters
  

