---
title: "Topo Data Visualization"
author: "Joey Hulbert"
date: "5/18/2022"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/redcedar/ServerFiles/redcedar/") 
```

```{r include=FALSE}
library(tidyverse)
library(randomForest)
library(caret)
library(rpart)
library(knitr)
```

# Purpose

The purpose of this markdown document is to list the steps we followed for visualization data with topo explanatory variables only.

# Data

The data used in the below models are described in the [Data Wrangle folder](./data-wrangle). 

## Response variables

**Note the below analysis uses the iNat data with 1510 observations. Amazing!**

* Response variables to consider
  + Number of unhealthy trees
  + Tree canopy symptoms

Below we visualizae data using canopy symptoms as the response variable

## Explanatory variables

* Explanatory variables included
  + Topo data
    + CTI - Climatic Moisture Index
    + SP - Slope Position
    + HLI - Heat Load Index
    + SDS - 2nd Derivitive slope
    + TRASP - Transformed Aspect - TRASP
    + IMI - Integrated Moisture Index
    + DEM - Elevation


  
```{r}
data <- read.csv("./data/observations-223649.csv")
#data.filtered <- data %>% select(c("id","field.tree.canopy.symptoms")) - remove all data except response variable of interest
```
  
```{r}
CTI <- read.csv("./data/topo_extract_files/CTI_extract_1500.csv")
CTI <- rename(CTI, CTI_VALUE = RASTERVALU)
CTI <- rename(CTI, id = ID2)
CTI <- CTI[c(3,7)]

HLI <- read.csv("./data/topo_extract_files/HLI_extract_1500.csv")
HLI <- rename(HLI, HLI_VALUE = RASTERVALU)
HLI <- rename(HLI, id = ID2)
HLI <- HLI[c(3,7)]

IMI <- read.csv("./data/topo_extract_files/IMI_extract_1500.csv")
IMI <- rename(IMI, IMI_VALUE = RASTERVALU)
IMI <- rename(IMI, id = ID2)
IMI <- IMI[c(3,7)]

SDS <- read.csv("./data/topo_extract_files/SDS_extract_1500.csv")
SDS <- rename(SDS, SDS_VALUE = RASTERVALU)
SDS <- rename(SDS, id = ID2)
SDS <- SDS[c(3,7)]

SP <- read.csv("./data/topo_extract_files/SP_extract_1500.csv")
SP <- rename(SP, SP_VALUE = RASTERVALU)
SP <- rename(SP, id = ID2)
SP <- SP[c(3,7)]

DEM <- read.csv("./data/topo_extract_files/DEM_extract_1500.csv")
DEM <- rename(DEM, DEM_VALUE = RASTERVALU)
DEM <- rename(DEM, id = ID2)
DEM <- DEM[c(3,7)]

TRASP <- read.csv("./data/topo_extract_files/TRASP_extract_1500.csv")
TRASP <- rename(TRASP, TRASP_VALUE = RASTERVALU)
TRASP <- rename(TRASP, id = ID2)
TRASP <- TRASP[c(3,7)]
```  

```{r}
combo <- left_join(data,CTI,by="id") %>% left_join(.,HLI,by="id") %>% left_join(.,IMI,by="id") %>% left_join(.,SDS,by="id") %>% left_join(.,SP,by="id") %>% left_join(.,DEM,by="id") %>% left_join(.,TRASP,by="id") 
```

```{r}
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combo <- combo %>% droplevels()
combo$field.tree.canopy.symptoms <- as.factor(combo$field.tree.canopy.symptoms)
#levels(combo$field.tree.canopy.symptoms)
```

# Data wrangling

There are multiple methods to group the response variables deepening on desired resolution or fineness of the model. 

* Response grouping options
  + Original (no change)
  + Filtered Symptoms
    + We could drop trees outside of main symptom classes
       + Note loss of data outside of categories
  + Reclassified symptoms
    + Here we can created an 'other' variable to represent trees with the below symptoms
        + extra cone crop
        + yellowing canopy
    + We can also group trees with 'old dead tops' and 'new dead tops' into a 'dead top' category
  + Binary (healthy/unhealthy)

For now, we can move forward with the binary response grouping because it is the broadest and presumably the easiest for the model to classify with. 


All tree health categories

```{r}
combo %>% group_by(field.tree.canopy.symptoms) %>% count()
```

## Filter data

### Filter observatoins

We also want to filter observations to only those within Oregon and Washington as those are the only observations we extracted TOPO data for.

```{r}
combo.filtered <- combo %>% filter(HLI_VALUE!="NA")
```

## Group response variables to binary values

```{r}
level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Unhealthy", "New Dead Top (red or brown needles still attached)" = "Unhealthy", "Old Dead Top (needles already gone)" = "Unhealthy", "Tree is dead" = "Unhealthy", "Multiple Symptoms (please list in Notes)" = "Unhealthy", "Extra Cone Crop" = "Unhealthy", "Browning Canopy" = "Unhealthy","Branch Dieback or 'Flagging'" = "Unhealthy", "Other (please describe in Notes)" = "Unhealthy", "Yellowing Canopy" = "Unhealthy")

binary <- combo.filtered
binary$field.tree.canopy.symptoms <- recode_factor(binary$field.tree.canopy.symptoms, !!!level_key)
#levels(binary$field.tree.canopy.symptoms)
binary <- binary %>% filter(field.tree.canopy.symptoms!="Candelabra top or very old spike top (old growth)") %>% droplevels()
binary$field.tree.canopy.symptoms <- as.factor(binary$field.tree.canopy.symptoms)
```

Binary tree health categories

```{r}
binary %>% group_by(field.tree.canopy.symptoms) %>% count()
```

## Data Visualization

```{r}
ggplot(binary,aes(HLI_VALUE,fill=field.tree.canopy.symptoms)) +geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw()
```


```{r}
ggplot(binary,aes(CTI_VALUE,fill=field.tree.canopy.symptoms)) +geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw()
```

```{r}
ggplot(binary,aes(SP_VALUE,fill=field.tree.canopy.symptoms)) +geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw()
```

```{r}
ggplot(binary,aes(TRASP_VALUE,fill=field.tree.canopy.symptoms)) +geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw()
```

```{r}
ggplot(binary,aes(IMI_VALUE,fill=field.tree.canopy.symptoms)) +geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw()
```

```{r}
ggplot(binary,aes(SDS_VALUE,fill=field.tree.canopy.symptoms)) +geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw()
```

```{r}
ggplot(binary,aes(DEM_VALUE,fill=field.tree.canopy.symptoms)) +geom_density(alpha=0.5) + scale_fill_manual(name="Tree Condition",values=c("#7fcdbb","#fe9929")) +theme_bw()
```

