---
title: "Analyses - Random Forests v3"
author: "Joey Hulbert"
date: "2/6/2022"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/redcedar/ServerFiles/redcedar/") 
```

```{r include=FALSE}
library(tidyverse)
library(randomForest)
library(caret)
library(rpart)
library(knitr)
```

```{r}
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
``` 

## Version 3 Notes

* Comments about this Analysis version
  + It does not subset the data into test or train
  + It only models 'filtered symptoms (5 groups)


### Purpose

The purpose of this analysis document is to ensure the reproducability of the results by guiding the reader through the random forest analysis of the factors associated with the health of western redcedar. 

### Approach

Root data were shared by citizen scientists in the [Wester Redcedar Dieback Map](https://www.inaturalist.org/projects/western-redcedar-dieback-map) project on iNaturalist. 

* These data were then combined with 
 + Climate data derived using the ClimateNA tool. Instructions for how this tool was used are included in the [climate subfolder](/climate).
 + Topographic data extraced from DEM rasters. Preliminary methods for extracting topo data from GPS points are available in the [topo subfolder](/topo). 
 + Soils data downloaded from gSSURRGO databases for Washington and Oregon. Instructions for extracting the soils data using Arc GIS Pro are included  in the [soils subfolder](/soils). 

All of the data used in the below analyses are described in the [Data Wrangle folder](./data-wrangle). 


### Limitations

* Below are some limitations to these results to clarify
  + representation of the data were heavily biased to the western side of Washington and Oregon (west of the Cascades). 


### Next steps
 * Do following before next commit
    + calculate better topographic variables (aspect, and those that combine both topo and climate data?) 
    + look into what is happening with -9999 values in components
    + does it make sense to group dead tops?
    + try hyperparameter tuning (see https://www.youtube.com/watch?v=Ptd2NXdtHl4 and https://jsimkins2.github.io/geog473-673/random-forest-modeling.html) 


### Notes from meeting with MC
  * Next steps
    + Correlation model between all variables - variables will be highly correlated 
    + Pick top ten that are uncorrelated
      + knock out some of the variables, keep the uncorrelated variables
      + Then see how model preforms
    + explore validatoin tools built into randomforests that could be used instead of subsetting the data
    + if you're not predicting into the future, than we may not need to split the dataset
  
Just interested in identifying drivers

### Data

The data used in the below visualizations are described in the [Data Wrangle folder](./data-wrangle). 

```{r}
data <- read.csv("./data/observations-9.19.21.csv")
```

```{r}
gps <- data[c(1,22,23)]
gps <- rename(gps,lat = latitude) %>% `colnames<-`(c("ID2","lat","long")) %>% mutate(el = ".") #columns were ranamed to match format for ClimateNA tool.
```

```{r}
normals <- read.csv("./data/gps_Normal_1991_2020MSY.csv")
colnames(normals) <- str_c("norm_",colnames(normals)) #change column names - not to be confused with decadal data
normals <- rename(normals, id = norm_ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
lastdecade <- read.csv("./data/gps_Decade_2011_2020MSY.csv")
lastdecade <- rename(lastdecade, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
topo <- read.csv("./data/gps plus DEM and TOPO data.csv")
topo <- rename(topo, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
wasoils <- read.csv("./data/Washington_GPS_Soils_Data_muaggatt_component.csv")
orsoils <- read.csv("./data/Oregon_GPS_Soils_Data_muagget_component.csv") #note misspelling of muaggatt
wasoils <- rename(wasoils, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
orsoils <- rename(orsoils, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
pnwsoils <- bind_rows(wasoils,orsoils)
```

```{r}
#combo.normals <- left_join(data,normals,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id")
#combo.decade <- left_join(data,lastdecade,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id") 
combo <- left_join(data,lastdecade,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id")  %>% left_join(.,normals,by="id") 
```

```{r}
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combo <- combo %>% droplevels()
combo$field.tree.canopy.symptoms <- as.factor(combo$field.tree.canopy.symptoms)
#levels(combo$field.tree.canopy.symptoms)
```

```{r}
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
combo <- combo %>% droplevels()
combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
#levels(combo$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
```

```{r}
combo$field.optional...tree.size[combo$field.optional...tree.size == "Large"] <- "Large (too big to wrap arms around trunk)"
combo$field.optional...tree.size[combo$field.optional...tree.size == "Medium"] <- "Medium (can wrap arms around trunk)"
combo$field.optional...tree.size[combo$field.optional...tree.size == "Small"] <- "Small (can wrap hands around trunk)"
combo$field.optional...tree.size[combo$field.optional...tree.size == "Very Large"] <- "Very Large (would take many people to wrap arms around trunk)"
combo <- combo %>% droplevels()
combo$field.optional...tree.size <- as.factor(combo$field.optional...tree.size)
#levels(combo$field.optional...tree.size)
```



#### Filter Data

Filter trees to only those with soils data (Oregon and Washington)

```{r}
combo <- combo %>% filter(MAP >0 & RASTERVALU>=0, na.rm=TRUE)
```

All tree health categories

```{r}
combo %>% group_by(field.tree.canopy.symptoms) %>% count()
```

Here we filter data to only include those from health categories with decent number of observations

```{r}
combo.filtered.symptoms <- combo %>% filter(field.tree.canopy.symptoms=="Healthy"|field.tree.canopy.symptoms=="Thinning Canopy"|field.tree.canopy.symptoms=="New Dead Top (red or brown needles still attached)"|field.tree.canopy.symptoms=="Old Dead Top (needles already gone)"|field.tree.canopy.symptoms=="Tree is dead") %>% droplevels()
#levels(filtered.symptoms$field.tree.canopy.symptoms)
```

```{r}
combo.filtered.symptoms %>% group_by(field.tree.canopy.symptoms) %>% count()
```


We also need to filter the data to only include response and explanatory variables we're interested in. For example, whether a sound clip was included in the iNat data is not important. 

We also need to remove other response variables like "field.percent.canopy.affected...." so it is not used as a predictor for tree health. 

> Note it might be interesting to know if the user was an important factor in predicting if the tree is healthy/unhealthy. 

There are also a number of factors that should probably be removed because they may be biasing the data. For example, only trees with the 'other factor' question may only be answered for unhealthy trees. We need to think about this a bit more. 

```{r}
combo.filtered <- combo.filtered.symptoms %>% select(-c("field.optional...did.the.tree.have.heat.damage"
,"field.percent.canopy.affected...."
,"field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight."
,"field.percent.of.trees..of.same.species..within.sight.that.are.unhealthy"
,"muaggatt_mukey"
,"component_mukey"
,"component_cokey"
,"id"  
,"observed_on_string"                                                                     
,"observed_on"                                                                            
,"time_observed_at"                                                                       
,"time_zone"                                                                              
,"user_id"                                                                                
,"user_login"                                                                             
,"created_at"                                                                             
,"updated_at"
,"quality_grade"                                                                          
,"license"                                                                                
,"url"                                                                                    
,"image_url"                                                                              
,"sound_url"                                                                              
,"tag_list"                                                                               
,"description"                                                                            
,"num_identification_agreements"                                                          
,"num_identification_disagreements"                                                       
,"captive_cultivated"                                                                     
,"oauth_application_id"                                                                   
,"place_guess"                                                                            
,"latitude"                                                                               
,"longitude"                                                                              
,"positional_accuracy"                                                                    
,"private_place_guess"                                                                    
,"private_latitude"                                                                       
,"private_longitude"                                                                      
,"public_positional_accuracy"                                                             
,"geoprivacy"                                                                             
,"taxon_geoprivacy"                                                                       
,"coordinates_obscured"                                                                   
,"positioning_method"                                                                     
,"positioning_device"                                                                     
,"place_town_name"                                                                        
,"place_county_name"                                                                      
,"place_state_name"                                                                       
,"place_country_name"                                                                     
,"place_admin1_name"                                                                      
,"place_admin2_name"                                                                      
,"species_guess"                                                                          
,"scientific_name"
,"common_name"                                                                            
,"iconic_taxon_name"                                                                      
,"taxon_id"          
,"field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage."
,"field.optional...what..other.factors..were.observed." 
, "field.optional...were.there.any.other.unhealthy.plant.species.on.the.site."             
 ,"field.optional...timing.of.symptoms.estimate"                                           
, "field.optional...estimated.time.spent.to.make.this.observation....of.minutes."          
, "field.optional...can.we.follow.up.with.you."
,"field_1.x"                                                                              
, "lat.x"                                                                                  
,"long.x"                                                                                 
, "el.x"      
, "OID_"                                                                                   
, "field_1.y"                                                                              
, "lat.y"                                                                                  
, "long.y"                                                                                 
, "el.y"                                                                                   
, "Aspect.y"                                                                               
, "Profilecurv.y"                                                                          
, "Tangentialc.y"                                                                          
, "Slope.y"                                                                                
, "PNWSRTMDEM.y"                                                                           
, "RASTERVALU"                                                                             
, "VAT_MapunitRaster_10m_Count"                                                            
, "VAT_MapunitRaster_10m_MUKEY"
,"X" 
,"Latitude"                                    
,"Longitude"
,"field.notes"
,"norm_X"                                                                                   
,"norm_Latitude"                                                                            
,"norm_Longitude"                                                                           
,"norm_Elevation"
))
# combo.filtered <- combo.filtered[c(45,50:59,68:332,337:341,355:394,396:503,510:774)] 
```

Remove variables with variables that have near zero standard deviations (entire column is same value)

```{r include=FALSE}
#nearZeroVar(combo.filtered)
combo.filtered.nearzerovar <- combo.filtered[,-nearZeroVar(combo.filtered)]
#write.csv(combo.decade.filtered,file="./data/troubleshoot.csv")
```

```{r}
combo.filtered.nearzerovar <-as.data.frame(unclass(combo.filtered.nearzerovar),stringsAsFactors=TRUE) #change all chars to factors
```

#### Impute data

We continue to get the below error, but were able to work around it by imputing the data. 

> Error in randomForest.default(m, y, ...) : Need at least two classes to do classification.

To impute the data we have to remove factors with >53 levels. 

The below code lists the number of levels for the variables that are factors. 

```{r echo=TRUE, include=FALSE}
for (n in names(combo.filtered.nearzerovar))
  if (is.factor(combo.filtered.nearzerovar[[n]])) {
    print(n)
    print(length(levels(combo.filtered.nearzerovar[[n]])))
  }
```

* The following factors had more than 53 levels
  + "muaggatt_musym" 
  + "muaggatt_muname"
  + "component_compname"
  + "component_geomdesc"
  + "component_taxclname"
  + "component_taxsubgrp"

```{r}
less.fifty.three <- combo.filtered.nearzerovar %>% select(-c("muaggatt_musym","muaggatt_muname","component_compname","component_geomdesc","component_taxclname","component_taxsubgrp")) # again, this may seem like a more difficult way than subsetting with df[-c(240,241,279,299,322,326)], but its easier in the long run when column numbers change as the data is filtered or added to. 
```

Imputed data table

```{r}
data.imputed <- rfImpute(field.tree.canopy.symptoms ~ ., data= less.fifty.three, iter=6)
```
### Model

```{r}
rf <- randomForest(field.tree.canopy.symptoms ~ ., data=data.imputed, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
```


```{r}
rf
```

> Note the 'out of bag' error rate of 38.5%


```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=data.imputed$field.tree.canopy.symptoms)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.35) +
  theme_bw() + 
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```

```{r}
importance <- varImp(rf,scale=TRUE)
plot(importance)
```


Important variables 

