---
title: "Topo Data Analysis"
author: "Joey Hulbert"
date: "5/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/redcedar/ServerFiles/redcedar/") 
```

```{r include=FALSE}
library(tidyverse)
library(randomForest)
library(caret)
library(rpart)
library(knitr)
```

# Purpose

The purpose of this markdown document is to list the steps we followed for refining the models using topo data only.

# Data

The data used in the below models are described in the [Data Wrangle folder](./data-wrangle). 

## Response variables

**Note the below analysis uses the iNat data with 1510 observations. Amazing!**

* Response variables to consider
  + Number of unhealthy trees
  + Tree canopy symptoms

Below we compare models using canopy symptoms as the response variable

## Explanatory variables

* Explanatory variables included
  + Topo data
    + CTI - Climatic Moisture Index
    + SP - Slope Position
    + HLI - Heat Load Index
    + SDS - 2nd Derivitive slope
    + TRASP - Transformed Aspect - TRASP
    + IMI - Integrated Moisture Index
    + DEM - Elevation
  
Do we want to include topo explanatory variables from iNat data? 

  * Possible additional explanatory variables
    + iNat data
      + tree size
      + slope position
      + site description
  
```{r}
data <- read.csv("./data/observations-223649.csv")
#data.filtered <- data %>% select(c("id","field.tree.canopy.symptoms")) - remove all data except response variable of interest
```
  
```{r}
CTI <- read.csv("./data/topo_extract_files/CTI_extract_1500.csv")
CTI <- rename(CTI, CTI_VALUE = RASTERVALU)
CTI <- rename(CTI, id = ID2)
CTI <- CTI[c(3,7)]

HLI <- read.csv("./data/topo_extract_files/HLI_extract_1500.csv")
HLI <- rename(HLI, HLI_VALUE = RASTERVALU)
HLI <- rename(HLI, id = ID2)
HLI <- HLI[c(3,7)]

IMI <- read.csv("./data/topo_extract_files/IMI_extract_1500.csv")
IMI <- rename(IMI, IMI_VALUE = RASTERVALU)
IMI <- rename(IMI, id = ID2)
IMI <- IMI[c(3,7)]

SDS <- read.csv("./data/topo_extract_files/SDS_extract_1500.csv")
SDS <- rename(SDS, SDS_VALUE = RASTERVALU)
SDS <- rename(SDS, id = ID2)
SDS <- SDS[c(3,7)]

SP <- read.csv("./data/topo_extract_files/SP_extract_1500.csv")
SP <- rename(SP, SP_VALUE = RASTERVALU)
SP <- rename(SP, id = ID2)
SP <- SP[c(3,7)]

DEM <- read.csv("./data/topo_extract_files/DEM_extract_1500.csv")
DEM <- rename(DEM, DEM_VALUE = RASTERVALU)
DEM <- rename(DEM, id = ID2)
DEM <- DEM[c(3,7)]

TRASP <- read.csv("./data/topo_extract_files/TRASP_extract_1500.csv")
TRASP <- rename(TRASP, TRASP_VALUE = RASTERVALU)
TRASP <- rename(TRASP, id = ID2)
TRASP <- TRASP[c(3,7)]
```  

```{r}
combo <- left_join(data,CTI,by="id") %>% left_join(.,HLI,by="id") %>% left_join(.,IMI,by="id") %>% left_join(.,SDS,by="id") %>% left_join(.,SP,by="id") %>% left_join(.,DEM,by="id") %>% left_join(.,TRASP,by="id") 
```

```{r}
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combo$field.tree.canopy.symptoms[combo$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combo <- combo %>% droplevels()
combo$field.tree.canopy.symptoms <- as.factor(combo$field.tree.canopy.symptoms)
#levels(combo$field.tree.canopy.symptoms)
```

# Data wrangling

There are multiple methods to group the response variables deepening on desired resolution or fineness of the model. 

* Response grouping options
  + Original (no change)
  + Filtered Symptoms
    + We could drop trees outside of main symptom classes
       + Note loss of data outside of categories
  + Reclassified symptoms
    + Here we can created an 'other' variable to represent trees with the below symptoms
        + extra cone crop
        + yellowing canopy
    + We can also group trees with 'old dead tops' and 'new dead tops' into a 'dead top' category
  + Binary (healthy/unhealthy)

For now, we can move forward with the binary response grouping because it is the broadest and presumably the easiest for the model to classify with. 


All tree health categories

```{r}
combo %>% group_by(field.tree.canopy.symptoms) %>% count()
```

## Filter data

### Filter observatoins

We also want to filter observations to only those within Oregon and Washington as those are the only observations we extracted TOPO data for.

```{r echo=TRUE, include=FALSE}
colSums(is.na(combo.filtered))
```

```{r}
combo.filtered <- combo %>% filter(HLI_VALUE!="NA")
```



### Filter predictors

We also need to filter the data to only include response and explanatory variables we're interested in. For example, whether a sound clip was included in the iNat data is not important. 

We also need to remove other response variables like "field.percent.canopy.affected...." so it is not used as a predictor for tree health. 

> Note it might be interesting to know if the user was an important factor in predicting if the tree is healthy/unhealthy. 

There are also a number of factors that should probably be removed because they may be biasing the data. For example, only trees with the 'other factor' question may only be answered for unhealthy trees. We need to think about this a bit more. 

Remove all iNat factors not useful as explanatory variables in current analysis (e.g. observed_on)

```{r}
combo.filtered <- combo.filtered %>% select(-c("field.optional...site.type",
                                      "field.optional...site.hydrology",
                                      "field.optional...site.location.description",
                                      "field.optional...site.area.disturbance.level",
                                      "field.optional...tree.size",
  "field.optional...slope.position",
  "field.optional...did.the.tree.have.heat.damage"
,"field.percent.canopy.affected...."
,"field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight."
,"field.percent.of.trees..of.same.species..within.sight.that.are.unhealthy"
,"field.optional...access.to.water"
,"id"  
,"observed_on_string"                                                                     
,"observed_on"                                                                            
,"time_observed_at"                                                                       
,"time_zone"                                                                              
,"user_id"                                                                                
,"user_login"                                                                             
,"created_at"                                                                             
,"updated_at"
,"quality_grade"                                                                          
,"license"                                                                                
,"url"                                                                                    
,"image_url"                                                                              
,"sound_url"                                                                              
,"tag_list"                                                                               
,"description"                                                                            
,"num_identification_agreements"                                                          
,"num_identification_disagreements"                                                       
,"captive_cultivated"                                                                     
,"oauth_application_id"                                                                   
,"place_guess"                                                                            
,"latitude"                                                                               
,"longitude"                                                                              
,"positional_accuracy"                                                                    
,"private_place_guess"                                                                    
,"private_latitude"                                                                       
,"private_longitude"                                                                      
,"public_positional_accuracy"                                                             
,"geoprivacy"                                                                             
,"taxon_geoprivacy"                                                                       
,"coordinates_obscured"                                                                   
,"positioning_method"                                                                     
,"positioning_device"                                                                     
,"place_town_name"                                                                        
,"place_county_name"                                                                      
,"place_state_name"                                                                       
,"place_country_name"                                                                     
,"place_admin1_name"                                                                      
,"place_admin2_name"                                                                      
,"species_guess"                                                                          
,"scientific_name"
,"common_name"                                                                            
,"iconic_taxon_name"                                                                      
,"taxon_id"          
,"field.other.factors...are.there.signs.or.symptoms.of.insect..diseases..or.other.damage."
,"field.optional...what..other.factors..were.observed." 
, "field.optional...were.there.any.other.unhealthy.plant.species.on.the.site."             
 ,"field.optional...timing.of.symptoms.estimate"                                           
, "field.optional...estimated.time.spent.to.make.this.observation....of.minutes."          
, "field.optional...can.we.follow.up.with.you."
,"field.notes"
))
# combo.filtered <- combo.filtered[c(45,50:59,68:332,337:341,355:394,396:503,510:774)] 
```

## Group response variables to binary values

```{r}
level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Unhealthy", "New Dead Top (red or brown needles still attached)" = "Unhealthy", "Old Dead Top (needles already gone)" = "Unhealthy", "Tree is dead" = "Unhealthy", "Multiple Symptoms (please list in Notes)" = "Unhealthy", "Extra Cone Crop" = "Unhealthy", "Browning Canopy" = "Unhealthy","Branch Dieback or 'Flagging'" = "Unhealthy", "Other (please describe in Notes)" = "Unhealthy", "Yellowing Canopy" = "Unhealthy")

binary <- combo.filtered
binary$field.tree.canopy.symptoms <- recode_factor(binary$field.tree.canopy.symptoms, !!!level_key)
#levels(binary$field.tree.canopy.symptoms)
binary <- binary %>% filter(field.tree.canopy.symptoms!="Candelabra top or very old spike top (old growth)") %>% droplevels()
binary$field.tree.canopy.symptoms <- as.factor(binary$field.tree.canopy.symptoms)
```

Binary tree health categories

```{r}
binary %>% group_by(field.tree.canopy.symptoms) %>% count()
```


# Approach

# Models

* Random forest model criteria
  + number of splits was not defined to allow automatic selection,
  + ntree = 2001, importance = TRUE, proximity = TRUE, na.action = na.omit
  + seed was set for reproducability with: set.seed(71)
  
#### Full Model

```{r full rf model}
set.seed(71)
rf <- randomForest(field.tree.canopy.symptoms ~ ., data=binary, ntree=2001, importance=TRUE, na.action=na.omit, proximity=TRUE)
```

```{r}
rf
```

```{r}
plot(rf)
```

```{r fig.height=6,fig.width=9}
#Below code copied from: https://github.com/StatQuest/random_forest_demo/blob/master/random_forest_demo.R as described here: https://www.youtube.com/watch?v=6EXPYzbfLCE

## Start by converting the proximity matrix into a distance matrix. 

distance.matrix <- as.dist(1-rf$proximity)

mds.stuff <- cmdscale(distance.matrix, eig=TRUE, x.ret=TRUE)

## calculate the percentage of variation that each MDS axis accounts for...
mds.var.per <- round(mds.stuff$eig/sum(mds.stuff$eig)*100, 1)

## now make a fancy looking plot that shows the MDS axes and the variation:
mds.values <- mds.stuff$points
mds.data <- data.frame(Sample=rownames(mds.values),
  X=mds.values[,1],
  Y=mds.values[,2],
  Status=binary$field.tree.canopy.symptoms)

ggplot(data=mds.data, aes(x=X, y=Y, label=Sample)) + 
  geom_text(aes(color=Status)) +  
  stat_ellipse(geom="polygon",aes(fill=Status,color=Status),alpha=0.45) +
  theme_bw() + 
  scale_color_manual(values=c("#7fcdbb","#fe9929")) +
  scale_fill_manual(values=c("#7fcdbb","#fe9929")) +
  xlab(paste("MDS1 - ", mds.var.per[1], "%", sep="")) +
  ylab(paste("MDS2 - ", mds.var.per[2], "%", sep="")) +
  ggtitle("MDS plot using (1 - Random Forest Proximities)")

#By default, the stat_ellipse function draws a 95% confidence level for a multivariate t-distribution. You can modify this level with level argument.

#more info for ellipse https://r-charts.com/correlation/scatter-plot-ellipses-ggplot2/
```

```{r fig.height=9,fig.width=12}
varImpPlot(rf)
```
