---
title: "Data Visualization and Exploration v2"
author: "Joey Hulbert"
date: "1/1/2022"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/redcedar/ServerFiles/redcedar/") 
```


```{r include=FALSE}
#install.packages("tidyverse")
library(tidyverse)
```


# Purpose

The purpose of this document is to track and provide instructions for reproducing the analyses of the [Wester Redcedar Dieback Map](https://www.inaturalist.org/projects/western-redcedar-dieback-map) project data.

# Approach

The overall approach is to model empirical data collected by community scientists with ancillary environmental data (climate, topographic, and soils) to identify important predictors of western redcedar dieback.

# Open questions

The below bullets are questions or comments that should be resolved before final analyses

* Considerations
  + Should we filter data to remove GPS points with poor accuracy?
  + Data averaged over last decade or last 30 years?
  + Better approach to calculate topographic data?
  + Should we restrict data to Washington and Oregon because that is where soils data is available. 
  + Should we look at data from last decade only or also from last 30 years?

# Notes

  * Important notes to consider when visualizing these data.
    + **Observations are filtered to only include those with Soils Data**
    + Data is filtered to show most common symptoms only. 

# Data

The data used in the below visualizations are described in the [Data Wrangle folder](./data-wrangle). 

```{r}
data <- read.csv("./data/observations-9.19.21.csv")
```

```{r}
gps <- data[c(1,22,23)]
gps <- rename(gps,lat = latitude) %>% `colnames<-`(c("ID2","lat","long")) %>% mutate(el = ".") #columns were ranamed to match format for ClimateNA tool.
```

```{r}
normals <- read.csv("./data/gps_Normal_1991_2020MSY.csv")
normals <- rename(normals, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
lastdecade <- read.csv("./data/gps_Decade_2011_2020MSY.csv")
lastdecade <- rename(lastdecade, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
topo <- read.csv("./data/gps plus DEM and TOPO data.csv")
topo <- rename(topo, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
```

```{r}
wasoils <- read.csv("./data/Washington_GPS_Soils_Data_muaggatt_component.csv")
orsoils <- read.csv("./data/Oregon_GPS_Soils_Data_muagget_component.csv") #note misspelling of muaggatt
wasoils <- rename(wasoils, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
orsoils <- rename(orsoils, id = ID2) #rename column 'ID2' (changed for ClimateNA tool) back to 'id'
pnwsoils <- bind_rows(wasoils,orsoils)
```

```{r}
#combinednormals <- left_join(data,normals,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id")
combo.decade <- left_join(data,lastdecade,by="id")  %>% left_join(.,topo,by="id") %>% left_join(.,pnwsoils,by="id")
```

```{r}
combo.decade$field.tree.canopy.symptoms[combo.decade$field.tree.canopy.symptoms=="Multiple Symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo.decade$field.tree.canopy.symptoms[combo.decade$field.tree.canopy.symptoms=="multiple symptoms"] <-"Multiple Symptoms (please list in Notes)"
combo.decade$field.tree.canopy.symptoms[combo.decade$field.tree.canopy.symptoms=="thinning foliage"] <-"Thinning Canopy"
combo.decade$field.tree.canopy.symptoms[combo.decade$field.tree.canopy.symptoms=="healthy"] <-"Healthy"
combo.decade$field.tree.canopy.symptoms[combo.decade$field.tree.canopy.symptoms=="dead top"] <-"Old Dead Top (needles already gone)"
combo.decade <- combo.decade %>% droplevels()
combo.decade$field.tree.canopy.symptoms <- as.factor(combo.decade$field.tree.canopy.symptoms)
#levels(combo.decade$field.tree.canopy.symptoms)
```

```{r}
combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="4"] <- "4-6"
combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="5"] <- "4-6"
combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.[combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.=="2"] <- "2-3"
combo.decade <- combo.decade %>% droplevels()
combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight. <- as.factor(combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
#levels(combo.decade$field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)
```

```{r}
combo.decade$field.optional...tree.size[combo.decade$field.optional...tree.size == "Large"] <- "Large (too big to wrap arms around trunk)"
combo.decade$field.optional...tree.size[combo.decade$field.optional...tree.size == "Medium"] <- "Medium (can wrap arms around trunk)"
combo.decade$field.optional...tree.size[combo.decade$field.optional...tree.size == "Small"] <- "Small (can wrap hands around trunk)"
combo.decade$field.optional...tree.size[combo.decade$field.optional...tree.size == "Very Large"] <- "Very Large (would take many people to wrap arms around trunk)"
combo.decade <- combo.decade %>% droplevels()
combo.decade$field.optional...tree.size <- as.factor(combo.decade$field.optional...tree.size)
#levels(combo.decade$field.optional...tree.size)
```

```{r}
combo.decade <- combo.decade %>% filter(MAP >0 & RASTERVALU>=0, na.rm=TRUE)
```

```{r}
combo.decade %>% group_by(field.tree.canopy.symptoms) %>% count()
```

```{r}
filtered.symptoms <- combo.decade  %>% filter(field.tree.canopy.symptoms=="Healthy"|field.tree.canopy.symptoms=="Thinning Canopy"|field.tree.canopy.symptoms=="New Dead Top (red or brown needles still attached)"|field.tree.canopy.symptoms=="Old Dead Top (needles already gone)"|field.tree.canopy.symptoms=="Tree is dead") %>% droplevels()
#levels(filtered.symptoms$field.tree.canopy.symptoms)
```

```{r}
filtered.symptoms %>% group_by(field.tree.canopy.symptoms) %>% count()
```

```{r}
level_key <- c("Healthy" = "Healthy", "Thinning Canopy" = "Unhealthy", "New Dead Top (red or brown needles still attached)" = "Unhealthy", "Old Dead Top (needles already gone)" = "Unhealthy", "Tree is dead" = "Unhealthy", "Multiple Symptoms (please list in Notes)" = "Unhealthy", "Extra Cone Crop" = "Unhealthy", "Browning Canopy" = "Unhealthy","Branch Dieback or 'Flagging'" = "Unhealthy", "Other (please describe in Notes)" = "Unhealthy", "Yellowing Canopy" = "Unhealthy")

binary <- combo.decade
binary$field.tree.canopy.symptoms <- recode_factor(binary$field.tree.canopy.symptoms, !!!level_key)
#levels(binary$field.tree.canopy.symptoms)
```

```{r}
binary %>% group_by(field.tree.canopy.symptoms) %>% count()
```

# Data Visualization
### Climate
#### Decade 2011-2020 {.tabset .tabset-pills}
##### MAT

###### Symptoms

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

###### Site Severity

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot()+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="Decade 2011-2020",x="Symptoms")
```

###### Size Class

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.optional...tree.size,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="Decade 2011-2020",x="Symptoms")
```

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+facet_wrap(~field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

```{r include=FALSE}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +facet_wrap(~field.optional...tree.size) +labs(title="Decade 2011-2020",x="Symptoms")
```

##### MAP

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MAP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

##### AHM

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,AHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

##### SHM

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

###### Filtered SHM <200
```{r}
filtered.symptoms.shm <- filtered.symptoms %>% filter(SHM<200) %>% droplevels()
```

```{r}
ggplot(filtered.symptoms.shm,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms", y="Summer Heat-Moisture Index")
```

##### MSP

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

##### CMD

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```


##### Eref

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,Eref,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

***


### TOPO Data {.tabset .tabset-pills}
#### Aspect

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,Aspect.x,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Aspect (degrees)",x="Symptoms")
```

#### Slope

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,Slope.x,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Slope (degrees)",x="Symptoms")
```

#### Elevation

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,PNWSRTMDEM.x,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Elevation (meters)",x="Symptoms")
```

### Topo & Climate  {.tabset .tabset-pills}
##### Slope
###### CMD
```{r}
ggplot(filtered.symptoms,aes(Slope.x,CMD,color=field.tree.canopy.symptoms))+geom_point()
```

```{r}
ggplot(filtered.symptoms,aes(Slope.x,CMD,color=field.tree.canopy.symptoms))+geom_smooth()
```

```{r}
ggplot(filtered.symptoms,aes(Slope.x,CMD,color=field.tree.canopy.symptoms))+geom_smooth()+facet_wrap(~field.tree.canopy.symptoms)
```

##### Elevation
```{r}
ggplot(filtered.symptoms,aes(PNWSRTMDEM.x,CMD,color=field.tree.canopy.symptoms))+geom_smooth()
```

***

### Soils Data {.tabset .tabset-pills}

#### Bedrock Depth

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,muaggatt_brockdepmin,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

#### drclasscd


```{r}
ggplot(filtered.symptoms,aes(muaggatt_drclassdcd,field.tree.canopy.symptoms,color=field.tree.canopy.symptoms))+geom_count(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```
```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,fill=field.tree.canopy.symptoms))+geom_bar(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")+facet_wrap(~muaggatt_drclassdcd)
```

#### drclasswettest
```{r}
ggplot(filtered.symptoms,aes(muaggatt_drclasswettest,field.tree.canopy.symptoms,color=field.tree.canopy.symptoms))+geom_count(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,fill=field.tree.canopy.symptoms))+geom_bar(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")+facet_wrap(~muaggatt_drclasswettest)
```

#### hydgrpdcd
```{r}
ggplot(filtered.symptoms,aes(muaggatt_hydgrpdcd,field.tree.canopy.symptoms,color=field.tree.canopy.symptoms))+geom_count(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,fill=field.tree.canopy.symptoms))+geom_bar(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")+facet_wrap(~muaggatt_hydgrpdcd)
```

#### slopegradwta
```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,muaggatt_slopegradwta,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

#### niccdcdpct
```{r}
ggplot(filtered.symptoms,aes(field.tree.canopy.symptoms,muaggatt_niccdcdpct,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

***

## Binary Data {.tabset .tabset-pills}

### Climate {.tabset .tabset-pills}

#### MAT

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

##### Site Severity

```{r include=FALSE}
ggplot(binary,aes(field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot()+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="Decade 2011-2020",x="Symptoms")
```

##### Size Class

```{r include=FALSE}
ggplot(binary,aes(field.optional...tree.size,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(fill="Unhealthy Trees")+labs(title="Decade 2011-2020",x="Symptoms")
```

```{r include=FALSE}
ggplot(binary,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+facet_wrap(~field.number.of.additional.unhealthy.trees..of.same.species..in.area..within.sight.)+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

```{r include=FALSE}
ggplot(binary,aes(field.tree.canopy.symptoms,MAT,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +facet_wrap(~field.optional...tree.size) +labs(title="Decade 2011-2020",x="Symptoms")
```

#### MAP

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,MAP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### AHM

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,AHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### SHM

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

##### Filtered SHM <200
```{r}
binary.shm <- binary %>% filter(SHM<200) %>% droplevels()
```

```{r}
ggplot(binary.shm,aes(field.tree.canopy.symptoms,SHM,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### MSP

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,MSP,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### CMD

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,CMD,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

#### Eref

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,Eref,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Decade 2011-2020",x="Symptoms")
```

### Topo

```{r}
ggplot(binary,aes(PNWSRTMDEM.x,CMD,color=field.tree.canopy.symptoms))+geom_smooth()
```

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,Slope.x,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(title="Slope (degrees)",x="Symptoms")
```

### Soils {.tabset .tabset-pills}

#### Bedrock Depth

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,muaggatt_brockdepmin,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

#### drclasscd


```{r}
ggplot(binary,aes(muaggatt_drclassdcd,field.tree.canopy.symptoms,color=field.tree.canopy.symptoms))+geom_count(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```
```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,fill=field.tree.canopy.symptoms))+geom_bar(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")+facet_wrap(~muaggatt_drclassdcd)
```

#### drclasswettest
```{r}
ggplot(binary,aes(muaggatt_drclasswettest,field.tree.canopy.symptoms,color=field.tree.canopy.symptoms))+geom_count(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,fill=field.tree.canopy.symptoms))+geom_bar(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")+facet_wrap(~muaggatt_drclasswettest)
```

#### hydgrpdcd
```{r}
ggplot(binary,aes(muaggatt_hydgrpdcd,field.tree.canopy.symptoms,color=field.tree.canopy.symptoms))+geom_count(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,fill=field.tree.canopy.symptoms))+geom_bar(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")+facet_wrap(~muaggatt_hydgrpdcd)
```

#### slopegradwta
```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,muaggatt_slopegradwta,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```

#### niccdcdpct
```{r}
ggplot(binary,aes(field.tree.canopy.symptoms,muaggatt_niccdcdpct,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+coord_flip()+theme_bw() +labs(x="Symptoms")
```


#### component_taxgrtgroup
```{r}
ggplot(binary,aes(component_taxgrtgroup,field.tree.canopy.symptoms))+geom_count(show.legend=FALSE)+coord_flip()+theme_bw()
```
#### component_taxpartsize
```{r}
levels(binary$component_taxpartsize)
ggplot(binary,aes(component_taxpartsize,field.tree.canopy.symptoms))+geom_count(show.legend=FALSE)+coord_flip()+theme_bw()
```

```{r}
ggplot(binary,aes(component_taxpartsize,fill=field.tree.canopy.symptoms))+geom_bar(position="dodge")+coord_flip()+theme_bw()
```

#### muaggatt_wtdepaprjunmin

Water Table Depth - April - June - Minimum
```{r}
str(binary$muaggatt_wtdepaprjunmin)
ggplot(binary,aes(muaggatt_wtdepaprjunmin,field.tree.canopy.symptoms,fill=field.tree.canopy.symptoms))+geom_boxplot(show.legend=FALSE)+theme_bw()
```


***
